Inequality in socio-emotional skills: a cross-cohort comparison

### 2020-03-14 codes/R/paper/seskills_jpube_rev.Rnw
update with external RIF estimates

\usepackage{authblk}
\title{Inequality in socioemotional skills: a cross-cohort comparison}
\author{Orazio Attanasio \hspace{10pt} Richard Blundell \\  Gabriella Conti \hspace{10pt} Giacomo Mason}
\affil{\emph{Institute for Fiscal Studies} \\ \emph{Department of Economics, University College London}}
\date{\today}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}

@@

  \begin{abstract}
  We examine changes in inequality in socio-emotional skills very early in life in two British cohorts born 30 years apart. We construct socio-emotional scales comparable across cohorts for both boys and girls, using two validated instruments for the measurement of child behaviour. We identify two dimensions of socio-emotional skills for each cohort: `internalising' and `externalising', related to the ability of children to focus their concentration and to engage in interpersonal activities, respectively. Using recent methodological advances in factor analysis, we establish comparability in the inequality of these early skills across cohorts, but not in their average level. We document for the first time that inequality in these early skills has increased across cohorts, especially for boys and at the bottom of the distribution. We also document changes in conditional skills gaps across cohorts. We find an increase in the socio-emotional skills gap in the younger cohort for children born to mothers with higher socio-economic status (education and employment), and to mothers who smoked during pregnancy. The increase in inequality in early socio-emotional skills is particularly pronounced for boys. On the other hand, we find a decline in the skills gradient for children without a father figure in the household. Lastly, we document that socio-emotional skills measured at a much earlier age than in most of the existing literature are significant predictors of outcomes both in adolescence and adulthood, in particular health and health behaviours. Our results show the importance of formally testing comparability of measurements to study skills differences across groups, and in general point to the role of inequalities in the early years for the accumulation of health and human capital across the life course.

  	Keywords: Inequality, Socio-emotional skills, Cohort studies, Measurement invariance

	JEL Classification: J13, J24, I14, I24, C38
  \end{abstract}
\vspace{20pt}

\scriptsize{Corresponding Author: Orazio Attanasio, Department of Economics, University College London,
Gordon Street, London WC1H0AX, UK. Email: o.attanasio@ucl.ac.uk. We thank participants to the NBER/LSE Trans-Atlantic Public Economics Seminar (TAPES) and our two discussants Hilary Hoynes and Gabriel Ulyssea for excellent comments. We're especially thankful to George Ploubidis for his guidance in the early stages of the project. This work is based on data from the British Cohort Study and the Millennium Cohort Study. We are grateful to The Centre for Longitudinal Studies, UCL Institute of Education, for the collection of the data, and to the UK Data Archive and UK Data Service for making them available. However, they bear no responsibility for the analysis or interpretation of the data. We would like to thank all cohort members and their families, who generously gave their time to participate in the surveys. This project has received funding from the European Research Council (ERC) under the European Union's Horizon 2020 research and innovation programme (grant agreement No 695300 - HKADeC - ERC-2015-AdG/ERC-2015-AdG and grant agreement No 819752 - DEVORHBIOSHIP - ERC-2018COG). We also thank the DynaHEALTH consortium (H2020 - 633595 DynaHEALTH) for funding and support. GM is funded by an Economic and Social Research Council scholarship.}
\end{titlepage}

\input{text/paper_jpube_1introlit.tex}
\input{text/paper_jpube_2datamethods.tex}
\input{text/paper_jpube_3resultsconclusion.tex}

@@

<<READ, cache=FALSE >>=
library(here)
read_chunk(here("codes/R/1_load.R"))
read_chunk(here("codes/R/2_efa.R"))
read_chunk(here("codes/R/2b_efa_all.R"))
read_chunk(here("codes/R/3_inv_main.R"))
read_chunk(here("codes/R/3_inv_aux.R"))
read_chunk(here("codes/R/3_inv_no511.R"))
read_chunk(here("codes/R/3_inv_fit.R"))
read_chunk(here("codes/R/4_bootstrap.R"))
read_chunk(here("codes/R/5_selectvars.R"))
read_chunk(here("codes/R/5_sumstats.R"))
read_chunk(here("codes/R/6_graphs.R"))
read_chunk(here("codes/R/6_graphs_ineq.R"))
read_chunk(here("codes/R/6_graphs_item.R"))
read_chunk(here("codes/R/7_ineq_quantiles.R"))
read_chunk(here("codes/R/8_regs_dep.R"))
read_chunk(here("codes/R/8_regs_outc.R"))
read_chunk(here("codes/R/9_rif.R"))
read_chunk(here("codes/R/99_auxfun.R"))

@@

<<RIF_PLOTS_PREP, dependson="SELECTDATA">>=
# make RIF decomposition plots
@

@@

@
\justify {\scriptsize \emph{Notes}: The table shows the mean values of harmonised variables (and the standard deviation, for continuous ones). The \emph{estimation sample} is the subsample used in the analysis. The \emph{full sample} is the entire sample of non-multiparous children in both cohorts, residing in England at birth. Mean estimates for the full sample in the MCS cohort are weighted to account for survey design.
\par}
\end{table}

@@

@@

@
\justify {\scriptsize \emph{Notes}: The table shows differences between quantiles of the distribution of socioemotional skills, by gender and cohort. Bootstrap confidence intervals with 1,000 repetitions are in brackets. The distribution is a factor score obtained from the factor model in \autoref{sec:methods}. These distributions are shown in \autoref{fig:dens}.
\par}
\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{landscape}
\begin{table}
\caption{Determinants of Socioemotional Skills}\label{tab:determ}
\centering\scriptsize
<<REGS_ALL, results='asis', dependson="REGS_DEP">>=
library(texreg)

@@

@
\justify {\scriptsize \emph{Notes}: The table shows coefficients from linear regressions of children's socioemotional skills at five years of age on family background information. The dependent variable is a factor score obtained from the factor model in \autoref{sec:methods}. Col. (1) and (2) show coefficients and standard errors in parentheses, for male children in the BCS and MCS cohorts separately. The latter are obtained using 1,000 bootstrap repetitions, taking into account the factor estimation stage that precedes the regression. Col. (3) shows the p-value of a test that the coefficient is the same in the two cohorts. Col. (4) to (6) repeat for female children. Col. (7) to (12) repeat for internalising skills. All estimates additionally control for region of birth, mother height, number of previous stillbirths at child's birth, preterm birth, a dummy for missing gestational age, and number of other children in the household at child age 5. See \autoref{tab:harmvar} for a description of the variables used.
\par}
\end{table}
\end{landscape}

@@

@
\justify {\scriptsize \emph{Notes}: The table shows coefficients from linear regressions of cohort members' adolescent outcomes on their externalising and internalising socioemotional skills at five years of age. Col. (1) shows the mean of the outcome for males. Col. (2) regresses the outcome on the scores obtained from the factor model in \autoref{sec:methods}. Col. (3) additionally controls for cognitive ability at age five. This is a simple factor score obtained by aggregating the available cognitive measures. All standard errors in parentheses are obtained using 1,000 bootstrap repetitions, taking into account the factor estimation stage that precedes the regression. Col. (4) to (6) repeat for female cohort members.  All estimates additionally control for region of birth, maternal education (5), maternal employment (5), father occupation (5), maternal background (age, height, nonwhite ethnicity, number of children in HH), pregnancy (firstborn child, number of previous stillbirths, mother smoked in pregnancy, preterm birth, (log) birth weight). See \autoref{tab:harmvar} for a description of the variables used.
\par}
\end{table}

@@

@
\justify {\scriptsize \emph{Notes}: The table shows coefficients from linear regressions of BCS cohort members' adult outcomes on their externalising and internalising socioemotional skills at five years of age. Col. (1) shows the mean of the outcome for males. Col. (2) regresses the outcome on the scores obtained from the factor model in \autoref{sec:methods}. Col. (3) additionally controls for cognitive ability at age five. This is a simple factor score obtained by aggregating the available cognitive measures. All standard errors in parentheses are obtained using 1,000 bootstrap repetitions, taking into account the factor estimation stage that precedes the regression. Col. (4) to (6) repeat for female cohort members.  All estimates additionally control for region of birth, maternal education (5), maternal employment (5), father occupation (5), maternal background (age, height, nonwhite ethnicity, number of children in HH), pregnancy (firstborn child, number of previous stillbirths, mother smoked in pregnancy, preterm birth, (log) birth weight). See \autoref{tab:harmvar} for a description of the variables used.
\par}
\end{table}

@@

\begin{figure}
<<FACDENS, dependson="SELECTDATA", dev='tikz', fig.height=10, fig.width=10, out.width='\\linewidth'>>=
@
\caption{Distribution of factor scores}\label{fig:dens}
\justify {\scriptsize \emph{Notes}: The figure shows the distribution of the externalising and internalising socioemotional skills scores at age five obtained from the factor model, by gender and cohort. The scores are estimated from parameter estimates in \autoref{tab:measpars}, using an Empirical Bayes Modal approach. Higher scores correspond to \emph{better} skills. The distribution is estimated nonparametrically, using an Epanechnikov kernel. The figure also reports the $p$ value from Kolmogorov-Smirnov tests of equality between the distribution in BCS and MCS.
\par}
\end{figure}

\begin{figure}
<<FACINEQ_PS, dependson="SELECTDATA", dev='tikz', fig.height=10, fig.width=10, out.width='\\linewidth'>>=
fi <- fi_ps # select result
gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]],
                        fi$ineqlist[[3]],fi$ineqlist[[4]],
                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
@
\caption{Skill inequality by mother's education}\label{fig:ineqps}
\justify {\scriptsize \emph{Notes}: The figure shows unconditional mean values of socioemotional skills scores by gender, cohort, and mother's education at age five. Mother's education is a dummy for whether the mother continued schooling past the minimum leaving age, based on her date of birth. The four panels on top present mean and 95\% confidence intervals. Given that we cannot compare means of skills, all scores are normalised to take value zero for the `Compulsory' category, so that the gradient is emphasised. The bottom two panels present the unconditional distribution of mother's education.
\par}
\end{figure}

\begin{figure}
<<FACINEQ_SM, dependson="SELECTDATA", dev='tikz', fig.height=10, fig.width=10, out.width='\\linewidth'>>=
fi <- fi_sm # select result
gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]],
                        fi$ineqlist[[3]],fi$ineqlist[[4]],
                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
@
\caption{Skill inequality by mother's pregnancy smoking}\label{fig:ineqsm}
\justify {\scriptsize \emph{Notes}: The figure shows unconditional mean values of socioemotional skills scores by gender, cohort, and mother's pregnancy smoking. Mother's education is a dummy for whether the mother reported smoking during pregnancy. The four panels on top present mean and 95\% confidence intervals. Given that we cannot compare means of skills, all scores are normalised to take value zero for the `Non-smoker' category, so that the gradient is emphasised. The bottom two panels present the unconditional distribution of mother's pregnancy smoking
\par}
\end{figure}

\begin{figure}
<<FACINEQ_FSC, dependson="SELECTDATA", dev='tikz', fig.height=10, fig.width=10, out.width='\\linewidth'>>=
fi <- fi_fsc # select result
gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]],
                        fi$ineqlist[[3]],fi$ineqlist[[4]],
                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
@
\caption{Skill inequality by father's occupation}\label{fig:ineqfsc}
\justify {\scriptsize \emph{Notes}: The figure shows unconditional mean values of socioemotional skills scores by gender, cohort, and father's occupation at age five. Father's occupation is based on Registrar General's social class, with classes I to III Non Manual being `White collar' and classes III Manual to V (plus `other') being `Blue collar'. `No father figure' is defined as absence of a male figure living in the household. The four panels on top present mean and 95\% confidence intervals. Given that we cannot compare means of skills, all scores are normalised to take value zero for the `Blue collar' category, so that the gradient is emphasised. The bottom two panels present the unconditional distribution of father's occupation.
\par}
\end{figure}

\begin{figure}
<<RIF_PLOTS, dependson="RIF_PLOTS_PREP", dev='tikz', fig.height=8, fig.width=10, out.width='\\linewidth'>>=
gridExtra::grid.arrange(rif_plots[[1]],rif_plots[[2]], ncol=1, nrow=2)
@
\caption{RIF decomposition of quantile differences}\label{fig:rif}
\justify {\scriptsize \emph{Notes}: 
\par}
\end{figure}

@@

@@

\begin{table}
\caption{Loadings from exploratory factor analysis -- 11-item scale}\label{tab:load}
\centering\setstretch{1.2}\footnotesize
<<EFALOADTAB, dependson='MCSEFARES', results='asis'>>=
# combine results
library(kableExtra)
loadtab <- cbind(meantab$num, meantab$title, data.frame(cbind(bcsfa$loadings, mcsfa$loadings)))
colnames(loadtab) <- c("item", "title", "bcs1", "bcs2", "mcs1", "mcs2")

@@

@@

@
\justify {\scriptsize \emph{Notes}: The table displays the factor loadings obtained from exploratory factor analysis (EFA) on the full set of SDQ items from the BCS. Items denoted by * are used in the 11-item scale. The EFA is performed decomposing the polychoric correlation matrix of the items and using weighted least squares, and the solution is rescaled using oblique factor rotation (\emph{oblimin}). We use the \textbf{\textsf{R}} package \texttt{psych}, version 1.8.4 \citep{Revelle2018}.
\par}
\end{table}

@@

\begin{landscape}
\begin{table}[ht]
\caption{Measurement invariance fit comparison}\label{tab:fit}
\centering\setstretch{1.2} \scriptsize
<<FA_FIT_RES, dependson="FA_FIT", results='asis'>>=
require(xtable)

@@

@
\justify {\scriptsize \emph{Notes}: The table presents fit indices for models of different invariance levels, following \cite{Wu2016a} Col. (1) displays the number of estimated parameters for each model. Col. (2) and (8) present the value of the $\chi^2$ statistic and the pvalue of the test of equality with respect to the configural model. Col. (3)-(7) and (9)-(13) present alternative fit indices (AFIs), in absolute values and differences from the configural model respectively. RMSE = Root mean squared error of approximation; SRMR = standardised root mean residual; MFI = McDonald non-centrality index; CFI = comparative fit index; G-hat = gamma-hat. Panel A shows results for the whole sample of children in the BCS and MCS cohorts. Panel B is restricted to a subsample of children in the age range of maximum overlap between the two cohorts (59-61 months). Panel C shows results from a model excluding items 5 and 11 of the 11-item subscale.
\par}
\end{table}
\end{landscape}

@@

@
\justify {\scriptsize \emph{Notes}: The table presents fit indices for models of different invariance levels, following \cite{Wu2016a} Col. (1) displays the number of estimated parameters for each model. Col. (2) and (8) present the value of the $\chi^2$ statistic and the pvalue of the test of equality with respect to the configural model. Col. (3)-(7) and (9)-(13) present alternative fit indices (AFIs), in absolute values and differences from the configural model respectively. RMSE = Root mean squared error of approximation; SRMR = standardised root mean residual; MFI = McDonald non-centrality index; CFI = comparative fit index; G-hat = gamma-hat.
\par}
\end{table}
\end{landscape}

@@

>>>>

\usepackage{authblk}
\title{Inequality in socio-emotional skills: a cross-cohort comparison}
\author{Orazio Attanasio \hspace{10pt} Richard Blundell \\  Gabriella Conti \hspace{10pt} Giacomo Mason}
%\affil{\emph{Institute for Fiscal Studies} \\ \emph{Department of Economics, University College London}}
\date{22 February 2020}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}

@@

\begin{abstract}
We examine changes in inequality in socio-emotional skills very early in life in two British cohorts born 30 years apart. We construct comparable scales using two validated instruments for the measurement of child behaviour and identify two dimensions of socio-emotional skills: `internalising' and `eternalising'. Using recent methodological advances in factor analysis, we establish comparability in the inequality of these early skills across cohorts, but not in their average level. We document for the first time that inequality in socio-emotional skills has increased across cohorts, especially for boys and at the bottom of the distribution. We also formally decompose the sources of the increase in inequality and find that compositional changes explain half of the rise in inequality in externalising skills. On the other hand, the increase in inequality in internalising skills seems entirely driven by changes in returns to background characteristics. Lastly, we document that socio-emotional skills measured at an earlier age than in most of the existing literature are significant predictors of health and health behaviours. Our results show the importance of formally testing comparability of measurements to study skills differences across groups, and in general point to the role of inequalities in the early years for the accumulation of health and human capital across the life course.\\
 Keywords: Inequality, Socio-emotional skills, Cohort studies, Measurement invariance
JEL Classification: J13, J24, I14, I24, C38
\end{abstract}
\vspace{20pt}
\scriptsize{Corresponding Author: Gabriella Conti, Department of Economics, University College London, Gordon Street, London WC1H0AX, UK. Email: gabriella.conti@ucl.ac.uk. We thank participants to the NBER/LSE Trans-Atlantic Public Economics Seminar (TAPES) and our two discussants Hilary Hoynes and Gabriel Ulyssea for excellent comments. We're especially thankful to George Ploubidis for his guidance in the early stages of the project. This work is based on data from the British Cohort Study and the Millennium Cohort Study. We are grateful to The Centre for Longitudinal Studies, UCL Institute of Education, for the collection of the data, and to the UK Data Archive and UK Data Service for making them available. However, they bear no responsibility for the analysis or interpretation of the data. We would like to thank all cohort members and their families, who generously gave their time to participate in the surveys. This project has received funding from the European Research Council (ERC) under the European Union's Horizon 2020 research and innovation programme (grant agreement No 695300 - HKADeC - ERC-2015-AdG and grant agreement No 819752 - DEVORHBIOSHIP - ERC-2018COG). We also thank the DynaHEALTH consortium (H2020 - 633595 DynaHEALTH) for funding and support. GM was funded by an Economic and Social Research Council scholarship.}
\end{titlepage}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\input{text/paper_jpube_1introlit.tex}
\input{text/paper_jpube_2datamethods.tex}
\input{text/paper_jpube_3resultsconclusion.tex}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

<<READ, cache=FALSE >>=
library(here)
read_chunk(here("codes/R/1_load.R"))
read_chunk(here("codes/R/2_efa.R"))
read_chunk(here("codes/R/2b_efa_all.R"))
read_chunk(here("codes/R/3_inv_main.R"))
read_chunk(here("codes/R/3_inv_aux.R"))
read_chunk(here("codes/R/3_inv_no511.R"))
read_chunk(here("codes/R/3_inv_fit.R"))
read_chunk(here("codes/R/4_bootstrap.R"))
read_chunk(here("codes/R/5_selectvars.R"))
read_chunk(here("codes/R/5_sumstats.R"))
read_chunk(here("codes/R/6_graphs.R"))
read_chunk(here("codes/R/6_graphs_ineq.R"))
read_chunk(here("codes/R/6_graphs_item.R"))
read_chunk(here("codes/R/7_ineq_quantiles.R"))
read_chunk(here("codes/R/8_regs_dep.R"))
read_chunk(here("codes/R/8_regs_outc.R"))
read_chunk(here("codes/R/99_auxfun.R"))

@@

@@

@
\justify {\scriptsize \emph{Notes}: The table shows the mean values of harmonised variables (and the standard deviation, for continuous ones). The \emph{estimation sample} is the subsample used in the analysis. The \emph{full sample} is the entire sample of children in both cohorts, residing in England at birth. Mean estimates for the full sample in the MCS cohort are weighted to account for survey design.
\par}
\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

@
\justify {\scriptsize \emph{Notes}: The table shows differences between quantiles of the distributions of socio-emotional skills, by gender and cohort. Bootstrap confidence intervals with 1,000 repetitions are in brackets. The factor scores for socio-emotional skills are estimated using an empirical Bayes modal approach, using the parameter estimates from the factor model in Table \ref{tab:measpars}. These distributions are shown in \autoref{fig:dens}.
\par}
\end{table}

\begin{landscape}
\begin{table}
\caption{Determinants of Socioemotional Skills}\label{tab:determ}
\centering\scriptsize
<<REGS_ALL, results='asis', dependson="REGS_DEP">>=
library(texreg)

@@

@
\justify {\scriptsize \emph{Notes}: The table shows coefficients from linear regressions of children's socio-emotional skills at five years of age on family background characteristics. The dependent variable is a factor score obtained from the factor model in \autoref{sec:methods}. Col. (1) and (2) show coefficients and standard errors in parentheses, for male children in the BCS and MCS cohorts separately. The latter are obtained using 1,000 bootstrap repetitions, taking into account the factor estimation stage that precedes the regression. Col. (3) shows the p-value of a test that the coefficient is the same in the two cohorts. Col. (4) to (6) repeat for female children. Col. (7) to (12) repeat for internalising skills. All estimates additionally control for region of birth, mother height, number of previous stillbirths at child's birth, preterm birth, a dummy for missing gestational age, and number of other children in the household at child age 5. See \autoref{tab:harmvar} for a description of the variables used. $^{***}$p$\leq$0.01, $^{**}$p$\leq$0.05, $^{*}$p$\leq$0.1. 
\par}
\end{table}
\end{landscape}

@@

@
\justify {\scriptsize \emph{Notes}: The table shows coefficients from linear regressions of cohort members' adolescent outcomes on their externalising and internalising socio-emotional skills at five years of age. Col. (1) shows the mean of the outcome for males. Col. (2) regresses the outcome on the scores obtained from the factor model in \autoref{sec:methods}. Col. (3) additionally controls for cognitive ability at age five. This is a simple factor score obtained by aggregating the available cognitive measures. All standard errors in parentheses are obtained using 1,000 bootstrap repetitions, taking into account the factor estimation stage that precedes the regression. Col. (4) to (6) repeat for female cohort members.  All estimates additionally control for region of birth, maternal education (5), maternal employment (5), father occupation (5), maternal background (age, height, nonwhite ethnicity, number of children in the household), pregnancy (firstborn child, number of previous stillbirths, mother smoked in pregnancy, preterm birth, (log) birth weight). See \autoref{tab:harmvar} for a description of the variables used.  $^{***}$p$\leq$0.01, $^{**}$p$\leq$0.05, $^{*}$p$\leq$0.1. 
\par}
\end{table}

@@

@
\justify {\scriptsize \emph{Notes}: The table shows coefficients from linear regressions of BCS cohort members' adult outcomes on their externalising and internalising socio-emotional skills at five years of age. Col. (1) shows the mean of the outcome for males. Col. (2) regresses the outcome on the scores obtained from the factor model in \autoref{sec:methods}. Col. (3) additionally controls for cognitive ability at age five. This is a simple factor score obtained by aggregating the available cognitive measures. All standard errors in parentheses are obtained using 1,000 bootstrap repetitions, taking into account the factor estimation stage that precedes the regression. Col. (4) to (6) repeat for female cohort members.  All estimates additionally control for region of birth, maternal education (5), maternal employment (5), father occupation (5), maternal background (age, height, nonwhite ethnicity, number of children in the household), pregnancy (firstborn child, number of previous stillbirths, mother smoked in pregnancy, preterm birth, (log) birth weight). See \autoref{tab:harmvar} for a description of the variables used.  $^{***}$p$\leq$0.01, $^{**}$p$\leq$0.05, $^{*}$p$\leq$0.1. 
\par}
\end{table}

@@

\begin{figure}
<<FACDENS, dependson="SELECTDATA", dev='tikz', fig.height=10, fig.width=10, out.width='\\linewidth'>>=
@
\caption{Distribution of factor scores}\label{fig:dens}
\justify {\scriptsize \emph{Notes}: The figure shows the distribution of the externalising and internalising socio-emotional skills scores at age five obtained from the factor model, by gender and cohort. The scores are estimated from the parameter estimates in \autoref{tab:measpars}, using an Empirical Bayes Modal approach. Higher scores correspond to \emph{better} skills. The distribution is estimated nonparametrically, using an Epanechnikov kernel. The figure also reports the $p$-value from Kolmogorov-Smirnov tests of equality between the distribution in BCS and MCS.
\par}
\end{figure}

\begin{figure}
<<FACINEQ_PS, dependson="SELECTDATA", dev='tikz', fig.height=10, fig.width=10, out.width='\\linewidth'>>=
fi <- fi_ps # select result
gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]],
                        fi$ineqlist[[3]],fi$ineqlist[[4]],
                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
@
\caption{Skill inequality by mother's education}\label{fig:ineqps}
\justify {\scriptsize \emph{Notes}: The figure shows unconditional mean values of socio-emotional skills scores by gender, cohort, and mother's education at age five. Mother's education is a dummy for whether the mother continued schooling past the minimum leaving age, based on her date of birth. The four panels on top present mean and 95\% confidence intervals. Given that we cannot compare means of skills, all scores are normalised to take value zero for the `Compulsory' category, so that the gradient is emphasised. The bottom two panels present the unconditional distribution of mother's education.
\par}
\end{figure}

\begin{figure}
<<FACINEQ_SM, dependson="SELECTDATA", dev='tikz', fig.height=10, fig.width=10, out.width='\\linewidth'>>=
fi <- fi_sm # select result
gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]],
                        fi$ineqlist[[3]],fi$ineqlist[[4]],
                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
@
\caption{Skill inequality by mother's pregnancy smoking}\label{fig:ineqsm}
\justify {\scriptsize \emph{Notes}: The figure shows unconditional mean values of socio-emotional skills scores by gender, cohort, and mother's pregnancy smoking. Maternal smoking is a dummy for whether the mother reported smoking during pregnancy. The four panels on top present mean and 95\% confidence intervals. Given that we cannot compare means of skills, all scores are normalised to take value zero for the `Non-smoker' category, so that the gradient is emphasised. The bottom two panels present the unconditional distribution of mother's pregnancy smoking
\par}
\end{figure}

\begin{figure}
<<FACINEQ_FSC, dependson="SELECTDATA", dev='tikz', fig.height=10, fig.width=10, out.width='\\linewidth'>>=
fi <- fi_fsc # select result
gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]],
                        fi$ineqlist[[3]],fi$ineqlist[[4]],
                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
@
\caption{Skill inequality by father's occupation}\label{fig:ineqfsc}
\justify {\scriptsize \emph{Notes}: The figure shows unconditional mean values of socio-emotional skills scores by gender, cohort, and father's occupation at age five. Father's occupation is based on the Registrar General's social class, with classes I to III Non Manual being `White collar' and classes III Manual to V (plus `other') being `Blue collar'. `No father figure' is defined as absence of a male figure living in the household. The four panels on top present mean and 95\% confidence intervals. Given that we cannot compare means of skills, all scores are normalised to take value zero for the `Blue collar' category, so that the gradient is emphasised. The bottom two panels present the unconditional distribution of father's occupation.
\par}
\end{figure}

\begin{figure}
<<RIF_PLOTS, dev='tikz', fig.height=8, fig.width=10, out.width='\\linewidth'>>=
rif_estimates <- read.csv(here("rif_estimates", "rif_m.csv")) %>%
  mutate(effect = factor(effect, levels = c("Total Difference", "Composition Effect", "Coefficient Effect"))) %>%
  mutate(statistic = factor(statistic, levels = c("DQ 90-10", "DQ 75-25", "DQ 50-10", "DQ 90-50", "D Variance")))

ggplot(rif_estimates, aes(fill=effect, y=value, x=statistic)) + 
    geom_bar(position="dodge", stat="identity") +
    geom_text(aes(label=significance), position=position_dodge(width=0.9)) +
    geom_hline(yintercept=0) +
  facet_wrap(~ skill, ncol = 1) +
  scale_fill_brewer(palette="OrRd") +

  theme(
    strip.background = element_blank(),
    strip.placement = "outside",
    axis.title.y = element_blank(),
    axis.title.x = element_blank(),
    legend.position = "bottom",
    legend.title = element_blank(),
    legend.spacing.x = unit(.3, 'cm'),
    legend.box.just = "center"
  )

@
\caption{RIF decomposition of changes in measures of inequality in socio-emotional skills - Males}\label{fig:rif}
\justify {\scriptsize \emph{Notes}: The figures show the total changes in five measures of inequality in socio-emotional skills between the BCS and the MCS, and decomposes them in composition and coefficient effects, following the RIF decomposition (with reweighting) of \citet{firpo2018decomposing}. The top figure presents the decompositions for the externalizing skills score, and the bottom figure for the internalizing skills score. The five inequality measures considered are the quantile differences 90-10, 75-25, 50-10, 90-50, and the variance. Full results are in \autoref{tab:RIF-EXT-M} and \autoref{tab:RIF-INT-M}. Bootstrapped standard errors over the entire procedure (500 replications) were used to compute the $p$-values. $^{***}$p$\leq$0.01, $^{**}$p$\leq$0.05, $^{*}$p$\leq$0.1.
\par}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{table}
\caption{Loadings from exploratory factor analysis with 2 factors -- 11-item scale}\label{tab:load}
\centering\setstretch{1.2}\footnotesize
<<EFALOADTAB, dependson='MCSEFARES', results='asis'>>=
# combine results
library(kableExtra)
loadtab <- cbind(meantab$num, meantab$title, data.frame(cbind(bcsfa$loadings, mcsfa$loadings)))
colnames(loadtab) <- c("item", "title", "bcs1", "bcs2", "mcs1", "mcs2")

@@

@
\justify {\scriptsize \emph{Notes}: The table displays the factor loadings obtained from exploratory factor analysis (EFA) on the full set of SDQ items from the MCS. Items denoted by * are used in the 11-item scale. The EFA is performed decomposing the polychoric correlation matrix of the items and using weighted least squares, and the solution is rescaled using oblique factor rotation (\emph{oblimin}). We use the \textbf{\textsf{R}} package \texttt{psych}, version 1.8.4 \citep{Revelle2018}.
\par}
\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@
\justify {\scriptsize \emph{Notes}: The table presents fit indices for models of different invariance levels, following \cite{Wu2016a} Col. (1) displays the number of estimated parameters for each model. Col. (2) and (8) present the value of the $\chi^2$ statistic and the $p$-value of the test of equality with respect to the configural model. Col. (3)-(7) and (9)-(13) present alternative fit indices (AFIs), in absolute values and differences from the configural model respectively. RMSE = Root mean squared error of approximation; SRMR = standardised root mean residual; MFI = McDonald non-centrality index; CFI = comparative fit index; G-hat = gamma-hat.
\par}
\end{table}
\end{landscape}

@@

\begin{landscape}
\begin{table}[ht]
\caption{Measurement invariance fit comparison -- two factors}\label{tab:fit}
\centering\setstretch{1.2} \scriptsize
<<FA_FIT_RES, dependson="FA_FIT", results='asis'>>=
require(xtable)

@@

@
\justify {\scriptsize \emph{Notes}: The table presents fit indices for models of different invariance levels, following \cite{Wu2016a} Col. (1) displays the number of estimated parameters for each model. Col. (2) and (8) present the value of the $\chi^2$ statistic and the $p$-value of the test of equality with respect to the configural model. Col. (3)-(7) and (9)-(13) present alternative fit indices (AFIs), in absolute values and differences from the configural model respectively. RMSE = Root mean squared error of approximation; SRMR = standardised root mean residual; MFI = McDonald non-centrality index; CFI = comparative fit index; G-hat = gamma-hat. Panel A shows results for the whole sample of children in the BCS and MCS cohorts. Panel B is restricted to a subsample of children in the age range of maximum overlap between the two cohorts (59-61 months). Panel C shows results from a model excluding items 5 and 11 of the 11-item subscale.
\par}
\end{table}
\end{landscape}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% RIF decomposition tables
\input{text/paper_jpube_riftables.tex}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
